---
title: Assignment 3 - One sample rates and paramter contrasts. Due November 15, 11:59pm 2020
author:
  - name: EPIB607 - Inferential Statistics
    affiliation: a
#  - name: Second Author
#    affiliation: a,b
address:
  - code: a
    address: Fall 2020, McGill University
#  - code: b
#    address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890
lead_author_surname: Bhatnagar
#doi: "https://sahirbhatnagar.com/EPIB607/"
abstract: |
  In this assignment you will practice one sample rates and parameter contrasts with linear regression. State all assumptions. Provide confidence intervals with appropriate units. Answers should be given in full sentences (DO NOT just provide the number). All graphs and calculations are to be completed in an R Markdown document using the provided template (from A1). You are free to choose any function from any package to complete the assignment. Concise answers will be rewarded. Be brief and to the point. Please submit the compiled pdf report to Crowdmark by November 15, 2020, 11:59pm. You need to save your answers to each question in separate pdf files. You also need to upload your code separately to Q5. See https://crowdmark.com/help/ for details.  
# Optional: Acknowledgements
#acknowledgements: |
#  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
#  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
#  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
#  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).
# Optional: One or more keywords
keywords:
  - Rates
  - Parameter contrasts
  - Regression
papersize: letter
fontsize: 12pt
# Optional: Force one-column layout, default is two-column
one_column: true
# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true
# Optional: Enable one-sided layout, default is two-sided
#one_sided: true
# Optional: Enable section numbering, default is unnumbered
numbersections: true
# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5
# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true
# Optional: Bibliography 
bibliography: pinp
# Optional: Enable a 'Draft' watermark on the document
watermark: false
#footer_contents: "Assignment 2 due October 11, 2020 by 11:59pm"
output: pinp::pinp
# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{YourPackage-vignetteentry}
  %\VignetteKeywords{YourPackage, r, anotherkeyword}
  %\VignettePackage{Due October 11, 2020}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---



```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
```




# (12 points, 2 each) Seismicity before and after hydraulic fracturing in the Horn River Basin, northeast British Columbia 

Consult the article _Investigation of regional seismicity before and after hydraulic fracturing in the Horn River Basin, northeast British Columbia_ by Farahbod et al. (2014). 

a) Using just the year 2006 data in the first four columns of Table 2, calculate separate event rates for the periods when hydraulic fracturing  (i) was and (ii) was not in operation.
Express each rate as the number of events per year, and accompany it with a 95% CI.

b) Repeat the calculations for the individual years 2007 to 2011.

c) Repeat the  calculations for the entire period 2006 to 2011. 

d) Store your results in a tidy `data.frame`.

e) Are you comfortable using Poisson-based CIs for the entire period 2006 to 2011? i.e. does it look like the rate in non-HF days is homogeneous over the 6 years, i.e., do the 6 CIs largely overlap each other?

f) Does it look like the rate in HF days is homogeneous over the 6 years, i.e., do the 6 CIs largely overlap each other?



# (28 points) Simulation study for confidence intervals of counts

The goal of this simulation study is to estimate the coverage probabilties of three different confidence intervals for a Poisson count. In class we saw at least three methods to calculate the confidence interval for a count: 1) normal approximation, 2) the exact method and 3) using the Poisson quantile `stats::qpois` function. 

a) (12 points) Simulate 1000 trials from a Poisson($\mu$ = 5) distribution using the `stats::rpois` function. For each of these trials, calculate the 95% confidence interval for the count using each of the three methods mentioned above. For each of the three methods, calculate the coverage probability, i.e., the percentage of intervals which contain the true mean $\mu$. Describe your findings and comment on the differences between methods in terms of coverage probability. 

b) (16 points) Repeat the simulation study in a), but this time, with increasing values of $\mu$. Visualize the coverage probabilities as a function of $\mu$ for each of the three methods. Describe your findings and comment on the differences between methods in terms of coverage probabilities as a function of $\mu$, i.e., what happens to the coverage probabilities when the expected number of counts increases? Explain.


```{r, eval=FALSE, echo=FALSE}
library(dplyr)
library(mosaic)

set.seed(1234)
(sims <- rbinom(n = 1000, size = 10, prob = 0.1))

sapply(sims, function(i){
  p <- i/10
  z_ci <- p + qnorm(c(0.025, 0.975)) * sqrt(p * (1-p) / 10)
  dplyr::between(0.2, z_ci[1], z_ci[2])
}) %>% mean

sapply(sims, function(i){
  p <- (i + 2) / (10 + 2)
  z_ci <- p + qnorm(c(0.025, 0.975)) * sqrt(p * (1-p) / (10 + 2))
  dplyr::between(0.2, z_ci[1],z_ci[2])
}) %>% mean

sapply(sims, function(i){
  tt <- stats::binom.test(x = i, n = 10)
  z_ci <- tt$conf.int
  dplyr::between(0.2, z_ci[1],z_ci[2])
}) %>% mean


sapply(sims, function(i){
  tt <- mosaic::binom.test(x = i, n = 10)
  
}) %>% mean


vecs <- c()
for(i in sims){
  tt <- mosaic::binom.test(i, 10, method = )
  z_ci <- tt$conf.int
  vecs <- c(dplyr::between(0.2, z_ci[1],z_ci[2]),vecs)
}
mean(vecs)


tt <- stats::prop.test(2,10)
tt <- stats::binom.test(2,10)
tt$conf.int[1]
tt$conf.int[2]
```


# (30 points, 5 each) Weather Data Analysis


```{r, eval=FALSE, echo=FALSE}
# pacman::p_load_gh("ropensci/weathercan")
library(tidyverse)
library(weathercan)

# get stations with lots of historical weather
head(stations)
stations %>%
  filter(prov %in% c("QC")) %>%
  filter(interval == "day") %>%
  filter(start <= 2002) %>%
  filter(end >= 2016) %>% 
  dplyr::arrange(start)


# download Danville station
qc <- weather_dl(station_ids = 5345, start = "1871-01-01", end = "2020-10-31", interval = "day")
qc <- qc[!is.na(qc$max_temp),]
qc_weather <- qc %>% 
  dplyr::select(station_name, station_id, prov, lat, lon, date, year, month, day, max_temp, min_temp, mean_temp) %>% 
  mutate(dayfromstart = 1:n()) %>% 
  dplyr::select(dayfromstart, everything()) %>% 
  mutate(year = as.numeric(as.character(year)),
         month = as.numeric(as.character(month)),
         day = as.numeric(as.character(day)))

readr::write_csv(qc_weather, file = "~/git_repositories/EPIB607/assignments/a3/qc_weather.csv")

qc_weather %>% 
  ggplot() + 
  geom_smooth(mapping = aes(x=dayfromstart, y = max_temp))


# stations %>% 
#   filter(prov == "QC") %>% 
#   pull(station_name) %>% 
#   table
# stations %>% 
#   filter(station_name %in% "MONTREAL/PIERRE ELLIOTT TRUDEAU INTL")

# yul <- weather_dl(station_ids = 30165, start = "2002-01-01", end = "2020-10-31")
# yul_weather <- yul %>% 
#   # dplyr::select(station_name, station_id, prov, lat)
#   group_by(year, month, day) %>% 
#   summarise(maxtemp = max(temp,na.rm=T),
#             meantemp = mean(temp,na.rm=T),
#             mintemp = min(temp, na.rm=T)) %>%
#   ungroup() %>% 
#   mutate(dayfromstart = 1:n()) %>% 
#   dplyr::select(dayfromstart, everything()) %>% 
#   filter(!is.na(meantemp))
```


Environment and Climate Change Canada (ECCC), formerly Environment Canada, is a federal department with the stated role of protecting the environment, conserving national natural heritage, and also providing weather and meteorological information. According to ECCC _warming over the 20th century is indisputable and largely due to human activities_ adding _Canada's rate of warming is about twice the global rate: a 2°C increase globally means a 3 to 4ºC increase for Canada_. Berkeley Earth has reported that [2015 was unambiguously the warmest year on record across the world](http://berkeleyearth.org/wp-content/uploads/2016/01/2015-Hottest-Year-BE-Press-Release-v1.0.pdf), with the Earth's temperature more than 1.0 C (1.8 F) above the 1850-1900 average.  
The data set `qc_weather.csv` contains daily temperature readings at [Danville weather station in Quebec](https://www.wunderground.com/weather/ca/danville) starting from June 6, 1880 (some dates are missing) that were scraped from the [Environment and Climate Change Canada (ECCC) website](https://climate.weather.gc.ca/historical_data/search_historic_data_e.html) using the [weathercan R package](https://docs.ropensci.org/weathercan/index.html). The daily measurements provided for you are:

- `year`: the year of the measurement.

- `month`: the month of the measurement; 1 = January, 2 = February,... , 12 = December.

- `day`: the day within the month (1-31).

- `max_temp`: the maximum recorded temperature that day.

- `mean_temp`: the average of recorded temperatures that day.

- `min_temp`: the minimum recorded temperature that day.

- `dayfromstart`: a number from 1 to 13501 that represents the number of days since June 5.  So 1 = June 6, 1880.


a) Create a new data set which contains all the rows that correspond to February in any year. Compare the `max_temp` in the February data between (1) those February days in 2015 and (2) February days in all other years using an appropriate figure and summary statistics. Describe your findings in 3 or fewer sentences.

<!--    \textcolor{NavyBlue}{The boxplot illustrates that temperatures in February 2018 were overall warmer than in other years. The mean maximum February temperature in 2018 is almost 9 degrees warmer than typical, and the median in 2018 is well above the third quartile in other years.}-->

```{r, echo=FALSE, eval=FALSE}


tt <- qc_weather %>% 
  filter(month == 2) %>%
  mutate(f2015 = ifelse(year==2015,TRUE,FALSE))

# tt %>% 
# ggplot(aes(x=f2015, y=max_temp)) + geom_boxplot() + facet_wrap(~factor(month))

# qc_weather$month %>% table

summary(lm(max_temp ~ f2015, data = tt))



```


b) Were those February days in 2015 warmer than February days in all other years ? Use an appropriate regression model to answer this question. Write the regression equation in terms of population parameters and define all parameters in your model. What is the parameter of interest?

c) Estimate the regression parameters using the data provided. Report the estimated coefficient for the parameter of interest and interpret it in the context of the problem.

d) Report the p-value for the parameter of interest. Based on the p-value, is there sufficient evidence at $\alpha = 0.05$ to support the claim that 2015 was the warmest year on record? What assumptions were used in calculating the p-value. 

e) Provide a potential explanation for why your analysis agrees or does not agree with the claim made by Berkeley Earth.

f) Use a permutation test to calculate the p-value for the parameter of interest and compare it with the one obtained in part d). Briefly discuss this comparison.

<!--    \textcolor{NavyBlue}{The mean and standard deviation of the replicates are 37.63 degrees and 1.79 degrees, respectively. Not a single one of the 10,000 replicates produced a sample mean as extreme as what was actually observed this past year.}-->

```{r, eval = FALSE, echo=FALSE}
#set the seed
set.seed(2018)

set.seed(101) ## for reproducibility
nsim <- 1000
res <- numeric(nsim) ## set aside space for results
for (i in 1:nsim) {
    ## standard approach: scramble response value
    perm <- sample(nrow(qc_weather))
    bdat <- transform(qc_weather, colonies=colonies[perm])
    ## compute & store difference in means; store the value
    res[i] <- mean(bdat[bdat$place=="field","colonies"])-
        mean(bdat[bdat$place=="forest","colonies"])
}
obs <- mean(ants[ants$place=="field","colonies"])-
    mean(ants[ants$place=="forest","colonies"])
## append the observed value to the list of results
res <- c(res,obs)
```





# (30 points, 6 each) Physical activity in NHANES

This problem uses data from the [National Health and Nutrition Examination Survey (NHANES)](https://cran.r-project.org/web/packages/NHANES/NHANES.pdf), a survey conducted annually by the US Centers for Disease Control (CDC). The dataset is available from the \texttt{NHANES} package.

Regular physical activity is important for maintaining a healthy weight, boosting mood, and reducing risk for diabetes, heart attack, and stroke. In this problem, you will be exploring the relationship between weight (\texttt{Weight}) and physical activity (\texttt{PhysActive}) using the NHANES data. Weight is measured in kilograms. The variable \texttt{PhysActive} is coded \texttt{Yes} if the participant does moderate or vigorous-intensity sports, fitness, or recreational activities, and \texttt{No} if otherwise. The objective is to compare weight between physically active and those who are not.

a) Explore the data.

    i. Identify how many individuals are physically active.
    
    ii. Create a plot that shows the association between weight and physical activity. Describe what you see.

b) Provide an appropriate regression model for the stated objective and state the parameter of interest. Give the regression equation in terms of population parameters and be sure to define each of the parameters in your model.
    
c) Fit a linear regression model to estimate the regression parameters. Report the estimated coefficients from the model and interpret each of them in the context of the problem.

d) Report a 95\% confidence interval for the parameter of interest and interpret the interval in the context of the problem. Based on the interval, is there sufficient evidence at $\alpha = 0.05$ to reject the null hypothesis of no association between weight and physical activity? State the assumptions used for calculating the 95\% confidence interval.

e) Provide a 95\% bootstrap confidence interval for the parameter of interest and compare it to the one in part d). Briefly discuss the comparison.



