---
title: DALITE Q3 - Parameters, Sampling Distributions and the Central Limit Theorem Solutions.

# Use letters for affiliations
author:
  - name: EPIB607 - Inferential Statistics
    affiliation: a
#  - name: Second Author
#    affiliation: a,b
address:
  - code: a
    address: Fall 2019, McGill University
#  - code: b
#    address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Bhatnagar and Hanley

# Place DOI URL or CRAN Package URL here
doi: "https://sahirbhatnagar.com/EPIB607/"

# Abstract
abstract: |
  This DALITE quiz will cover the building blocks of statistical inference. 

# Optional: Acknowledgements
#acknowledgements: |
#  This template package builds upon, and extends, the work of the excellent
#  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
#  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
#  gratefully acknowledged as this work would not have been possible without them.  Our extensions
#  are under the same respective licensing term
#  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
#  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).

# Optional: One or more keywords
keywords:
  - Parameters and statistics
  - Sampling distributions
  - Central Limit Theorem (CLT)

# Paper size for the document, values of letterpaper and a4paper
papersize: letter

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 12pt

# Optional: Force one-column layout, default is two-column
one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography 
bibliography: pinp

# Optional: Enable a 'Draft' watermark on the document
watermark: false

# Customize footer, eg by referencing the vignette
footer_contents: "DALITE Q3 due Sepetember 24, 2019 by 5pm"

# Produce a pinp document
output: pinp::pinp

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{YourPackage-vignetteentry}
  %\VignetteKeywords{YourPackage, r, anotherkeyword}
  %\VignettePackage{Due Sepetember 24, 2019}
  %\VignetteEngine{knitr::rmarkdown}
---

# Parameters Q1

Which of the following statements is false?

a) A parameter is a constant of unknown magnitude in a statistical model
b) The value of a parameter is in general unknown
c) A statistic is a number derived from a sample
d) **The population standard deviation can be estimated from the sample provided that we have a sample size greater than 30 (Correct)**


## Correct rationales

- No, you can have a standard deviation in a sample that has a 'n value' of smaller than 30 this would just change the estimated standard deviations accuracy.
- Population standard deviation can be estimated from a sample size regardless of the specific number


## Incorrect rationales

- Generally the sample standard deviation is estimated from the population standard deviation using the square root of the sample size.



# Sampling Distributions Q1

A newborn baby has extermely low birth weight (ELBW) if it weighs less than 1000 grams. A study of the health of such children in later years examined a random sample of 219 children. Their mean weight at birth was $\bar{y}$ = 810 grams. This sample mean ($\bar{y}$) is an unbiased estimator of the mean weight $\mu$ in the population of all ELBW babies. This means that

a) **In many samples from this population, the mean of many values of $\bar{y}$ will be equal to $\mu$ (Correct)**
b) As we take larger and larger samples from this population, the sample mean $\bar{y}$ will get closer and closer to $\mu$ 
c) In many samples from this population, the many values of the sample mean $\bar{y}$ will have a distribution that is close to Normal


## Correct rationales

- As the sample mean $\bar{y}$ is an unbiased estimator of the mean weight $\mu$ in the population, the mean of many values of $\bar{y}$ will be equal to $\mu$. Further, if we plot the sample means, the resulting sampling distribution of $\bar{y}$ will have the same mean value as the mean in the population distribution.
- Because the sample mean is an unbiased estimator of the mean weight. So the mean of many samples would be equal to $\mu$.

## Incorrect rationales

- The larger the samples from the population, the smaller the standard deviations and the closer the mean values are to the average of the population.



# CLT Q1

Cholesterol levels among fourteen-year-old boys are roughly Normal with mean 170 and standard deviation 30 milligrams per deciliter (mg/dl). You choose a simple random sample of 4 fourtheen-year-old boys and average their choleterol levels. If you do this many times, the mean of the average cholesterol levels you get will be close to

a) **170 (Correct)**
b) 170/4 = 42.5
c) 170/$\sqrt{4}$ = 85


## Correct rationales

- The population has a normal distribution, which means that the sample mean of n independent observations will also have a normal distribution with a mean equal to mu.


## Incorrect rationales




# CLT Q2

Cholesterol levels among fourteen-year-old boys are roughly Normal with mean 170 and standard deviation 30 milligrams per deciliter (mg/dl). You choose a simple random sample of 4 fourtheen-year-old boys and average their choleterol levels. If you do this many times, the standard deviation of the average cholesterol levels you get will be close to

a) 30
b) 4/$\sqrt{30}$ = 0.73
c) **30/$\sqrt{4}$ = 15 (Correct)**



## Correct rationales

- By the CLT, the sample cholesterol levels will be normally distributed with a standard deviation equal to the population standard deviation divided by the square root of the sample size.


## Incorrect rationales

- This is standard deviation we are discussing, not standard error, standard error requires we divide by square root of N
- Multiple samples will approximate the population parameter
- Taking multiple sample, sample mean will be closer to the population mean and sample standard deviation will be closer to population standard deviation.



# CLT Q3

The survival times of guinea pigs inoculated with an infections viral strain vary from animal to animal. The distribution of survival times is strongly skewed to the right. The central limit theorem says that

a) as we study more and more infected guinea pigs, their average survival time gets ever closer to the mean $\mu$ for all infected guinea pigs. 
b) the average survival time of a large number of infected guinea pigs has a distribution of the same shape (strongly skewed) as the distribution for individual infected guinea pigs
c) **the average survival time of a large number of infected guinea pigs has a distribution that is close to Normal. (Correct)**






## Correct rationales

- CLT states that as a sample becomes large enough, the distribution takes on a normal distribution despite the the distribution shape of its population.


## Incorrect rationales








<!-- 
We support several options via the YAML header

- Setting a DOI or URL footer, for example for the CRAN package URL,
  which is placed in the bottom-left footer of the title page and even pages;
- Setting a footer label, for example _YourPackage Vignette_ stating
  your package, which is placed in the bottom-right footer on odd
  pages;
- Setting a free-form author field used on the inside footer;
- Optional _Draft_ watermarking;



## References 

Here we differ from PNAS and suggest natbib. References will appear in
author-year form. Use `\citet{}`, `\citep{}`, etc as usual.

We default to the `jss.bst` style. To switch to a different bibliography
style, please use `biblio-style: style` in the YAML header.


## Inline R Code 

The PNAS sample included a fixed PNG image here, but this document prefers
to show the results and embedding of _R_ code. 

```{r figex, fig.width=3, fig.height=3, cache=TRUE, echo=TRUE, fig.cap="Narrow ggplot2 figure"}
library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) +
    geom_point(size=3, aes(colour=factor(cyl))) +
    theme(legend.position="none")
```

Here we use a standard knitr bloc with explicit options for

- figure width and height (`fig.width`, `fig.height`), both set to three inches;
- whether the code is shown (`echo=TRUE`); and
- the caption (`fig.cap`) as shown above.


## Digital Figures 

Markdown, Pandoc and LaTeX support `.eps` and `.pdf` files.

Figures and Tables should be labelled and referenced in the standard way
using the `\label{}` and `\ref{}` commands.

The R examples above show how to insert a column-wide
figure. To insert a figure wider than one column, please use the
`\begin{figure*}...\end{figure*}` environment. 

One (roundabout) way of doing this is to _not_ actually plot a figure, but to
save it in a file as the following segment shows:

```{r densityPlot, echo=TRUE}
library(ggplot2)
p <- ggplot(data = midwest,
            mapping = aes(x = area,
                          fill = state,
                          color = state)) +
    geom_density(alpha = 0.3)
## save to file
suppressMessages(ggsave("densities.pdf", p))
```

This file is then included via standard LaTeX commands.

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.66\textwidth, height=3.5in]{densities} 
  \end{center}
  \caption{Wide ggplot2 figure}\label{fig}
\end{figure*}


## Typeset Code (But Do Not Run It) 

We can also just show code.

```r
xx <- faithful[,"eruptions"]
fit <- density(xx)
plot(fit)
```

This simply used a pandoc bloc started and ended by three backticks,
with `r` as the language choice.  Similarly, _many_ other languages
can be typeset directly simply by relying on pandoc.


## Single column equations 

Authors may use 1- or 2-column equations in their article, according to
their preference.

To allow an equation to span both columns, options are to use the
`\begin{figure*}...\end{figure*}` environment mentioned above for
figures, or to use the `\begin{widetext}...\end{widetext}` environment
as shown in equation \ref{eqn:example} below.

Please note that this option may run into problems with floats and
footnotes, as mentioned in the [cuted package
documentation](http://texdoc.net/pkg/cuted). In the case of problems
with footnotes, it may be possible to correct the situation using
commands `\footnotemark` and `\footnotetext`.

\begin{equation}
  \begin{aligned}
(x+y)^3&=(x+y)(x+y)^2\\
       &=(x+y)(x^2+2xy+y^2) \\
       &=x^3+3x^2y+3xy^3+x^3. 
       \label{eqn:example} 
  \end{aligned}
\end{equation}


pandoc writes all tables using longtable, which fails in 2-column mode

  Species                    CBS     CV     G3
  ----------------------- ------ ------ ------
  1\. Acetaldehyde           0.0    0.0    0.0
  2\. Vinyl alcohol          9.1    9.6   13.5
  3\. Hydroxyethylidene     50.8   51.2   54.0

  : Comparison of the fitted potential energy surfaces and ab initio
  benchmark electronic energy calculations

-->

