---
title: Poll 1 - A Survey on Basic Statistics Concepts. 

# Use letters for affiliations
author:
  - name: EPIB607 - Inferential Statistics
    affiliation: a
#  - name: Second Author
#    affiliation: a,b
address:
  - code: a
    address: Fall 2018, McGill University
#  - code: b
#    address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Bhatnagar and Hanley

# Place DOI URL or CRAN Package URL here
doi: "https://sahirbhatnagar.com/EPIB607/"

# Abstract
abstract: |
  This live poll was conducted on the first day of class, September 6, 2018. Correct answers are indicated by check marks. Number of votes and number of participants are indicated in the figure legend. For some questions, several selections were allowed. 

# Optional: Acknowledgements
#acknowledgements: |
#  This template package builds upon, and extends, the work of the excellent
#  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
#  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
#  gratefully acknowledged as this work would not have been possible without them.  Our extensions
#  are under the same respective licensing term
#  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
#  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).

# Optional: One or more keywords
keywords:
  - Descriptive stats
  - Confidence Interval
  - p-value
  - Gaussian distribution
  - CLT
  - Simple linear regression

# Paper size for the document, values of letterpaper and a4paper
papersize: letter

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 10pt

# Optional: Force one-column layout, default is two-column
one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography 
bibliography: pinp

# Optional: Enable a 'Draft' watermark on the document
watermark: false

# Customize footer, eg by referencing the vignette
footer_contents: "Poll 1 - September 6, 2018"

# Produce a pinp document
output: pinp::pinp

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{YourPackage-vignetteentry}
  %\VignetteKeywords{YourPackage, r, anotherkeyword}
  %\VignettePackage{Due Sepetember 12, 2018}
  %\VignetteEngine{knitr::rmarkdown}
---


# Which of the following visual summaries do you know about? (select all that apply)

1. Histograms
2. Stem & Leaf plots
3. Boxplot
4. Dotplot
5. Scatterplot
6. Density plot
7. Violin plot


\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.15]{q1.jpg} 
  \end{center}
  \caption{267 votes - 69 participants}\label{fig}
\end{figure}

\newpage


# A study records the sex and weight (in kilograms) of 30 recently born bear cubs in Alaska. Which of the following statements is true? 

1. Both are categorical variables
2. Both are quantitative variables
3. Sex is categorical, Weight is quantitative 


\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.15]{q2.jpg} 
  \end{center}
  \caption{72 votes - 72 participants}
\end{figure}


# You are given a sequence of 5 numbers: 1, 5, 20, 35, 39. Which of the following statements are true? 

1. The median is equal to the mean
2. The median is less than the mean
3. The median is greater than the mean


\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.15]{q3.jpg} 
  \end{center}
  \caption{79 votes - 79 participants}
\end{figure}

\newpage

# In a distribution with a long right-tail, the median is

1. equal to the mean
2. less than the mean
3. greater than the mean

\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.15]{q4.jpg} 
  \end{center}
  \caption{77 votes - 77 participants}
\end{figure}


# If you add 7 to each value on a sequence of numbers, the standard deviation increases by 7

1. TRUE
2. FALSE




\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.15]{q5.jpg} 
  \end{center}
  \caption{77 votes - 77 participants}
\end{figure}


# If you double each value on a sequence of numbers, you double the standard deviation.

1. TRUE
2. FALSE


\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.15]{q6.jpg} 
  \end{center}
  \caption{75 votes - 75 participants}
\end{figure}


\newpage


# In a large set of observations, the distribution of observations follows the Gaussian curve quite closely

1. TRUE
2. FALSE



\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.15]{q7.jpg} 
  \end{center}
  \caption{77 votes - 77 participants}
\end{figure}




# A straight line relating a response variable y to an explanatory variable x is given by  y = a + bx. Which of the following statements are true? (select all that apply)

1. b is the intercept and a is the slope 
2. b is the slope and a is the intercept
3. the intercept a is the value of y when x=0

\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.15]{q8.jpg} 
  \end{center}
  \caption{119 votes - 79 participants}
\end{figure}

\newpage



# A p-value is (select all that apply)

1. conditional on the null hypothesis being true
2. a measure of evidence
3. the probability of the null hypothesis being true
4. a probability concerning the observed data
5. significant if it is less than 0.05
6. a probability that provides the same information as a confidence interval

\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.11]{q9.jpg} 
  \end{center}
  \caption{192 votes - 80 participants}
\end{figure}


# Which of the following statements are correct regarding a confidence interval? (select all that apply)

1. It often has the form: estimate +/- margin of error
2. Is an interval estimate computed from sample data that gives a range of plausible  values for a population parameter
3. The interval is constructed so that the value of the parameter will be captured between the endpoints of the interval with a chosen level of confidence
4. The confidence level is the success rate for the method
5. provides more useful information than the p-value

\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.11]{q10.jpg} 
  \end{center}
  \caption{186 votes - 79 participants}
\end{figure}


<!-- 
We support several options via the YAML header

- Setting a DOI or URL footer, for example for the CRAN package URL,
  which is placed in the bottom-left footer of the title page and even pages;
- Setting a footer label, for example _YourPackage Vignette_ stating
  your package, which is placed in the bottom-right footer on odd
  pages;
- Setting a free-form author field used on the inside footer;
- Optional _Draft_ watermarking;



## References 

Here we differ from PNAS and suggest natbib. References will appear in
author-year form. Use `\citet{}`, `\citep{}`, etc as usual.

We default to the `jss.bst` style. To switch to a different bibliography
style, please use `biblio-style: style` in the YAML header.


## Inline R Code 

The PNAS sample included a fixed PNG image here, but this document prefers
to show the results and embedding of _R_ code. 

```{r figex, fig.width=3, fig.height=3, cache=TRUE, echo=TRUE, fig.cap="Narrow ggplot2 figure"}
library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) +
    geom_point(size=3, aes(colour=factor(cyl))) +
    theme(legend.position="none")
```

Here we use a standard knitr bloc with explicit options for

- figure width and height (`fig.width`, `fig.height`), both set to three inches;
- whether the code is shown (`echo=TRUE`); and
- the caption (`fig.cap`) as shown above.


## Digital Figures 

Markdown, Pandoc and LaTeX support `.eps` and `.pdf` files.

Figures and Tables should be labelled and referenced in the standard way
using the `\label{}` and `\ref{}` commands.

The R examples above show how to insert a column-wide
figure. To insert a figure wider than one column, please use the
`\begin{figure*}...\end{figure*}` environment. 

One (roundabout) way of doing this is to _not_ actually plot a figure, but to
save it in a file as the following segment shows:

```{r densityPlot, echo=TRUE}
library(ggplot2)
p <- ggplot(data = midwest,
            mapping = aes(x = area,
                          fill = state,
                          color = state)) +
    geom_density(alpha = 0.3)
## save to file
suppressMessages(ggsave("densities.pdf", p))
```

This file is then included via standard LaTeX commands.

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.66\textwidth, height=3.5in]{densities} 
  \end{center}
  \caption{Wide ggplot2 figure}\label{fig}
\end{figure*}


## Typeset Code (But Do Not Run It) 

We can also just show code.

```r
xx <- faithful[,"eruptions"]
fit <- density(xx)
plot(fit)
```

This simply used a pandoc bloc started and ended by three backticks,
with `r` as the language choice.  Similarly, _many_ other languages
can be typeset directly simply by relying on pandoc.


## Single column equations 

Authors may use 1- or 2-column equations in their article, according to
their preference.

To allow an equation to span both columns, options are to use the
`\begin{figure*}...\end{figure*}` environment mentioned above for
figures, or to use the `\begin{widetext}...\end{widetext}` environment
as shown in equation \ref{eqn:example} below.

Please note that this option may run into problems with floats and
footnotes, as mentioned in the [cuted package
documentation](http://texdoc.net/pkg/cuted). In the case of problems
with footnotes, it may be possible to correct the situation using
commands `\footnotemark` and `\footnotetext`.

\begin{equation}
  \begin{aligned}
(x+y)^3&=(x+y)(x+y)^2\\
       &=(x+y)(x^2+2xy+y^2) \\
       &=x^3+3x^2y+3xy^3+x^3. 
       \label{eqn:example} 
  \end{aligned}
\end{equation}


pandoc writes all tables using longtable, which fails in 2-column mode

  Species                    CBS     CV     G3
  ----------------------- ------ ------ ------
  1\. Acetaldehyde           0.0    0.0    0.0
  2\. Vinyl alcohol          9.1    9.6   13.5
  3\. Hydroxyethylidene     50.8   51.2   54.0

  : Comparison of the fitted potential energy surfaces and ab initio
  benchmark electronic energy calculations

-->

