---
title: "In-class exercise on ROC curves"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Packages

```{r}
pacman::p_load(dplyr)
pacman::p_load(pROC)
```


# Survival of patients following admission to an adult intensive care unit (ICU)

The ICU study data set consists of a sample of 200 subjects who were part of a much larger study on survival of patients following admission to an adult intensive care unit (ICU). The major goal of this study was to develop a logistic regression model to predict the probability of survival to hospital discharge of these patients. A code sheet for the variables in this data is provided in `icu_codebase.pdf`.
The primary outcome variable is vital status at hospital discharge (`sta`). Clinicians associated with the study felt that a key determinant of survival was the type of admission, `type`.

```{r load}
load(here::here("inst/assignments/a8/icu.rda"))

# create binary version of sta
icu <- icu %>% 
  mutate(sta.int = as.numeric(sta) - 1)

head(icu)
```


# Intercept only


```{r}
# overall proportion hiv positive
round(sum(icu$sta.int)/nrow(icu),3)

# intercept-only logit model
fit0 <- glm(sta ~ 1, family=binomial(link=logit), data=icu)
print(summary(fit0), show.call=TRUE, digits=2, signif.stars = FALSE)

# checks
plogis(fit0$coefficients)
fitted(fit0)
icu$fit0fitted <- fitted(fit0)
```


# Several predictors

```{r}
fit1 <- glm(sta ~ type, 
            family=binomial(link=logit), 
            data=icu)
summary(fit1)
boxplot(fitted(fit1) ~ icu$sta)
icu$fit1fitted <- fitted(fit1)

fit2 <- glm(sta ~ type + age, 
            family=binomial(link=logit), 
            data=icu)

summary(fit2)

par(mfrow = c(1,2))
boxplot(fitted(fit1) ~ icu$sta)
boxplot(fitted(fit2) ~ icu$sta)
for(i in seq(0,1, by = 0.1)) abline(h = i, col = "red", lwd = 2)

icu$fit2fitted <- fitted(fit2)
```


# ROC Curves

```{r}
rocobj1 <- plot.roc(icu$sta.int, fitted(fit1),
                    percent=TRUE,
                    ci=TRUE, # compute AUC (of AUC by default)
                    print.auc=TRUE) # print the AUC (will contain the CI)

ciobj <- ci.se(rocobj1, # CI of sensitivity
               specificities=seq(0, 100, 5)) # over a select set of specificities
plot(ciobj, type="shape", col="#1c61b6AA") # plot as a blue shape
plot(ci(rocobj1, of="thresholds", thresholds="best")) # add one threshold
```



```{r}
rocobj2 <- plot.roc(icu$sta.int, fitted(fit2),
                    percent=TRUE,
                    ci=TRUE, # compute AUC (of AUC by default)
                    print.auc=TRUE) # print the AUC (will contain the CI)

ciobj2 <- ci.se(rocobj2, # CI of sensitivity
               specificities=seq(0, 100, 5)) # over a select set of specificities
plot(ciobj2, type="shape", col="#1c61b6AA") # plot as a blue shape
plot(ci(rocobj2, of="thresholds", thresholds="best")) # add one threshold
```

# Optimal cutoff

```{r}
boxplot(fitted(fit1) ~ icu$sta)
abline(h = coords(rocobj1, "best")$threshold)

boxplot(fitted(fit2) ~ icu$sta)
abline(h = coords(rocobj2, "best")$threshold[1])

best.thresh <- coords(rocobj2, "best")$threshold[1]
best.thresh

icu$model_based_class <- icu$fit2fitted > best.thresh
table(icu$model_based_class, icu$sta)
```



# Compare ROC curves


```{r}

plot.roc(icu$sta.int, fitted(fit1),
                    percent=TRUE,
                    ci=TRUE, # compute AUC (of AUC by default)
                    print.auc=TRUE) # print the AUC (will contain the CI)

plot.roc(icu$sta.int, fitted(fit2),
                    percent=TRUE,
                    ci=TRUE, # compute AUC (of AUC by default)
                    print.auc=TRUE, 
          col = "red", print.auc.y = 20,
         add = TRUE) # print the AUC (will contain the CI)

roc.test(rocobj1, rocobj2)





```



