\documentclass[letterpaper,12pt,twoside,]{pinp}

%% Some pieces required from the pandoc template
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

% pinp change: the geometry package layout settings need to be set here, not in pinp.cls
\geometry{layoutsize={0.95588\paperwidth,0.98864\paperheight},%
  layouthoffset=0.02206\paperwidth, layoutvoffset=0.00568\paperheight}

\definecolor{pinpblue}{HTML}{185FAF}  % imagecolorpicker on blue for new R logo
\definecolor{pnasbluetext}{RGB}{101,0,0} %


\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\title{Assignment 4 - Poisson and Logistic Regression. Due November 30, 11:59pm
2020}

\author[a]{EPIB607 - Inferential Statistics}

  \affil[a]{Fall 2020, McGill University}

\setcounter{secnumdepth}{5}

% Please give the surname of the lead author for the running footer
\leadauthor{Bhatnagar}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
 \keywords{  Rates |  Parameter contrasts |  Regression  }  

\begin{abstract}
In this assignment you will practice poisson and logistic regression.
State all assumptions. Provide confidence intervals with appropriate
units. Answers should be given in full sentences (DO NOT just provide
the number). All graphs and calculations are to be completed in an R
Markdown document using the provided template (from A1). You are free to
choose any function from any package to complete the assignment. Concise
answers will be rewarded. Be brief and to the point. Please submit the
compiled pdf report to Crowdmark by November 30, 2020, 11:59pm. You need
to save your answers to each question in separate pdf files. You also
need to upload your code separately to Q3. See
\url{https://crowdmark.com/help/} for details.
\end{abstract}

\dates{This version was compiled on \today} 

% initially we use doi so keep for backwards compatibility
% new name is doi_footer


\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.


\hypertarget{points-bcg-vaccination-and-leprosy}{%
\section{(50 points) BCG Vaccination and
Leprosy}\label{points-bcg-vaccination-and-leprosy}}

This question is based on the paper by \emph{Fine, P. E. M., et
al.~``Protective efficacy of BCG against leprosy in Northern Malawi.''
The Lancet 328.8505 (1986): 499-502.} The aim of the study was to
investigate whether BCG vaccination in early childhood, whose purpose is
to protect against leprosy, which is caused by a closely related
bacillus. New cases of leprosy reported during a given period in a
defined geographical area were examined for presence or absence of the
characteristic scar left by BCG vaccination. During approximately the
same period, a 100\% survey of the population of this area had been
carried out, and this survey included examination for BCG scar. The
tabulated data refer only to subjects under 35, because persons over the
age of 35 at the time of the study would have been children at a time
when vaccination was not widely available. The data in the table below
are from this unusual example of a case-control study in which the
controls were obtained from a 100\% cross-sectional survey of the study
base.

\begin{table}[h]
    \centering
    \begin{tabular}{rrcc}
        & \multicolumn{3}{c}{Case status} \\
        \cline{2-4}
        BCG scar & \ \hspace{3mm}\ & Leprosy cases & Healthy population survey (controls) \\
        \hline
        Present && 101 & 46,028 \\
        Absent && 159 & 34, 594 \\
        \hline
    \end{tabular}
\end{table}

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\item
  (8 points) Calculate \(\hat{p}_1\) and \(\hat{p}_2\), the two sample
  proportions of interest. Test the hypothesis for a difference in
  proportions; do the data suggest that the BCG vaccination in early
  childhood protects against leprosy? Be sure to check your assumptions.
\item
  (8 points) Formulate a hypothesis for a contingency-table approach to
  analyzing this data and test this hypothesis. Be sure to check your
  assumptions.
\item
  (8 points) Estimate the odds of BCG vaccination for leprosy cases and
  for the controls. Estimate the odds ratio, standard error (using
  Woolf's formula) and 95\% confidence interval and interpret it.
\item
  (8 points) Fit an appropriate logistic regression model and compare
  your results from part c). Do they agree?
\item
  (9 points) Summarize the conclusions from all three approaches
  (difference in proportions, contingency table, logistic regression).
  Do they agree? Explain.
\item
  (9 points) The following table shows the results of a
  computer-simulate study which picked 1000 controls at random. Re-run
  the logistic regression on these data and compare with part d).
  Comment on any differences or similarities (i.e.~the differences
  between using 80000 controls and 1000 controls).
\end{enumerate}

\begin{table}[h]
    \centering
    \begin{tabular}{rrcc}
        & \multicolumn{3}{c}{Case status} \\
        \cline{2-4}
        BCG scar & \ \hspace{3mm}\ & Leprosy cases & Healthy population survey (controls) \\
        \hline
        Present && 101 & 554 \\
        Absent && 159 & 446 \\
        \hline
    \end{tabular}
\end{table}

\hypertarget{points-population-mortality-rates-in-denmark}{%
\section{(50 points) Population mortality rates in
Denmark}\label{points-population-mortality-rates-in-denmark}}

The following table contains mortality data for males and females in
Denmark for 4 age groups over three time periods.

\vspace*{0.3in}

\begin{tabular}{l|l|r|r|r|r|r|r}
\hline
Year & Age & Female\_deaths & Female\_PT & Female\_rate & Male\_deaths & Male\_PT & Male\_rate\\
\hline
1980-1984 & 70-74 & 15989 & 586882.8 & 0.0272439 & 23810 & 456908.21 & 0.0521111\\
\hline
1980-1984 & 75-79 & 20838 & 454142.7 & 0.0458843 & 24707 & 300318.92 & 0.0822692\\
\hline
1980-1984 & 80-84 & 24073 & 297678.6 & 0.0808691 & 20319 & 167303.51 & 0.1214499\\
\hline
1980-1984 & 85-89 & 20216 & 147771.7 & 0.1368057 & 13524 & 74295.83 & 0.1820291\\
\hline
2000-2004 & 70-74 & 13912 & 521561.9 & 0.0266737 & 17360 & 436994.92 & 0.0397259\\
\hline
2000-2004 & 75-79 & 19731 & 471945.5 & 0.0418078 & 22477 & 341362.82 & 0.0658449\\
\hline
2000-2004 & 80-84 & 25541 & 369989.9 & 0.0690316 & 22992 & 217929.72 & 0.1055019\\
\hline
2000-2004 & 85-89 & 27135 & 226798.1 & 0.1196439 & 17444 & 104009.58 & 0.1677153\\
\hline
2005-2009 & 70-74 & 12179 & 540568.6 & 0.0225300 & 15782 & 472012.84 & 0.0334355\\
\hline
2005-2009 & 75-79 & 17273 & 444474.2 & 0.0388616 & 19547 & 344351.34 & 0.0567647\\
\hline
2005-2009 & 80-84 & 23513 & 363534.1 & 0.0646789 & 21781 & 230530.24 & 0.0944822\\
\hline
2005-2009 & 85-89 & 26842 & 237877.3 & 0.1128397 & 17811 & 114485.04 & 0.1555749\\
\hline
\end{tabular}

\begin{enumerate}
\def\labelenumi{\alph{enumi}.}
\tightlist
\item
  (15 points) Come up with a suitable regression model for this data.
  Write down the regression equation in terms of parameters and
  determinants.
\item
  (15 points) Estimate the parameters of this model using the data in
  the table above. Provide the fitted regression equation. The data is
  provided in the \texttt{denmark.csv} file in myCourses.
\item
  (10 points) Interpret the parameter for gender. Are mortality rates
  significantly different in males compared with females?
\item
  (10 points) Perform a goodness of fit test for the fitted model in
  part (b). Is this a good fit?
\end{enumerate}

%\showmatmethods


\bibliography{pinp}
\bibliographystyle{jss}



\end{document}

