---
title: Lab 005 - Inference for means
author:
  - name: EPIB607 - Inferential Statistics
    affiliation: a
#  - name: Second Author
#    affiliation: a,b
address:
  - code: a
    address: McGill University
#  - code: b
#    address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890
lead_author_surname: Bhatnagar and Hanley
doi: "https://sahirbhatnagar.com/EPIB607/"
abstract: |
  In this exercise you will practice calculating confidence intervals using the t-distribution and the bootstrap.
# Optional: Acknowledgements
#acknowledgements: |
#  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
#  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
#  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
#  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).
# Optional: One or more keywords
keywords:
  - Sampling distribution
  - Standard error
  - Normal distribution
  - Quantiles
  - Percentiles
  - Z-scores
papersize: letter
fontsize: 11pt
# Optional: Force one-column layout, default is two-column
one_column: true
# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true
# Optional: Enable one-sided layout, default is two-sided
#one_sided: true
# Optional: Enable section numbering, default is unnumbered
numbersections: true
# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5
# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true
# Optional: Bibliography 
bibliography: pinp
# Optional: Enable a 'Draft' watermark on the document
watermark: false
footer_contents: "in-class exercise on confidence intervals"
output: pinp::pinp
# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{YourPackage-vignetteentry}
  %\VignetteKeywords{YourPackage, r, anotherkeyword}
  %\VignettePackage{Due Sepetember 28, 2018}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, eval = FALSE, echo = FALSE)
```

```{r, eval=TRUE, echo=FALSE, inlcude=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(mosaic)
library(tidyr)
```



|R Code                     |       Value        |
|:--------------------------|:----------------------------------|
|`qnorm(p = c(0.05, 0.95))` |`r round(qnorm(p = c(0.05, 0.95)), 2)`    |
|`qnorm(p = c(0.025, 0.975))` |`r round(qnorm(p = c(0.025, 0.975)),2)`    |
|`qnorm(p = c(0.005, 0.995))` |`r round(qnorm(p = c(0.005, 0.995)),2)`    |
|`qt(p = c(0.025, 0.975), df = 400-1)` |`r round(qt(p = c(0.025, 0.975), df = 400-1), 2)`    |
|`qt(p = c(0.025, 0.975), df = 25-1)` |`r round(qt(p = c(0.025, 0.975), df = 25-1), 2)`    |
|`qt(p = c(0.025, 0.975), df = 20-1)` |`r round(qt(p = c(0.025, 0.975), df = 20-1), 2)`    |
|`qt(p = c(0.025, 0.975), df = 16-1)` |`r round(qt(p = c(0.025, 0.975), df = 16-1), 2)`    |



# Food intake and weight gain

If we increase our food intake, we generally gain weight. Nutrition scientists can calculate the amount of weight gain that would be associated with a given increase in calories. In one study, 16 nonobese adults, aged 25 to 36 years, were fed 1000 calories per day in excess of the calories needed to maintain a stable body weight. The subjects maintained this diet for 8 weeks, so they consumed a total of 56,000 extra calories. According to theory, 3500 extra calories will translate into a weight gain of 1 pound. Therfore we expect each of these subjects to gain 56,000/3500=16 pounds (lb). Here are the weights (given in the `weightgain.csv` file) before and after the 8-week period expressed in kilograms (kg):

```{r, eval=FALSE, echo=TRUE}
weight <- read.csv(here::here("inst/labs/005-one-sample-mean/weightgain.csv"))

```

<br>

a. Calculate a 95% confidence interval for the mean weight change and give a sentence explaining the meaning of the 95%. State your assumptions. 

```{r, echo=FALSE, eval=FALSE}
weight <- read.csv("~/git_repositories/EPIB607/exercises/inferencemeans/weightgain.csv")

# Creating new variable for weight change
weight$change <- weight$after-weight$before
weight$change_lb <- weight$change*2.2

# Calculating the mean of weight change and rounding
(ybar_change <- mean(weight$change))

# Calculating the sample standard deviation
(ssd_change <- sd(weight$change))

# sample size
(n <- nrow(weight))

# Calculating a 95% confidence interval version 1
qt_scaled <- function(p, df, mean, sd) {
  mean  + qt(p = p, df = df) * sd
}

(q1_ci95 <- qt_scaled(p = c(0.025, 0.975), 
                     df = nrow(weight) - 1, 
                     mean = ybar_change, 
                     sd = ssd_change / sqrt(n)))


# Calculating a 95% confidence interval version 2
ybar_change +  qt(p = c(0.025, 0.975), df = n - 1) * ssd_change / sqrt(n)

```


b. Calculate a 95% bootstrap confidence interval for the mean weight change and compare it to the one obtained in part (a). Comment on the bootstrap sampling distribution and compare it to the assumptions you made in part (a).

```{r, echo=FALSE, eval=FALSE, fig.asp=0.681}
q1_dist <- do(10000) * mean( ~ change, data = resample(weight))
hist(q1_dist$mean, col = "lightblue", lwd = 2)
round(quantile(~ mean, data = q1_dist, probs = c(0.001, 0.005, .025, 0.05, 
                                                 0.90, 0.95, 0.975, 0.99)),2)
```



c. Convert the units of the mean weight gain and 95% confidence interval to pounds. Note that 1 kilogram is equal to 2.2 pounds. Test the null hypothesis that the mean weight gain is 16 lbs. State your assumptions and justify your choice of test. Be sure to specify the null and alternative hypotheses. What do you conclude?


```{r, echo=FALSE, eval=FALSE, fig.asp=0.681}
round(quantile(~ mean, data = q1_dist, probs = c(.025, 0.975)) * 2.2,2)
```



\newpage

# Attitudes toward school

The Survey of Study Habits and Attitudes (SSHA) is a psychological test that measures the motivation, attitude toward school, and study habits of students. Scores range from 0 to 200. The mean score for U.S. college students is about 115, and the standard deviation is about 30. A teacher who suspects that older students have better attitudes toward school gives the SSHA to 25 students who are at least 30 years of age. Their mean score is $\bar{y}$ = 132.2 with a sample standard deviation $s = 28$. 

a. The teacher asks you to carry out a formal statistical test for her hypothesis. Perform a test, provide a 95% confidence interval and state your conclusion clearly. 
b. What assumptions did you use in part (a). Which of these assumptions is most important to the validity of your conclusion in part (a).



# Does a full moon affect behavior?

Many people believe that the moon influences the actions of some individuals. A study of dementia patients in nursing homes recorded various types of disruptive behaviors every day for 12 weeks. Days were classified as moon days if they were in a 3-day period centered at the day of the full moon. For each patient, the average number of disruptive behaviors was computed for moon days and for all other days. The hypothesis is that moon days will lead to more disruptive behavior. We look at a data set consisting of observations on 15 dementia patients in nursing homes (available in the `fullmoon.csv` file): 

```{r, echo=TRUE, eval=FALSE}
fullmoon <- read.csv("fullmoon.csv")
```

```{r, eval=FALSE, echo=FALSE}
# moon <- load("~/Downloads/FullMoon.RData")
# FullMoon$aggdiff <- NULL
# colnames(FullMoon) <- c("patient","moon_days", "other_days")
# write.csv(FullMoon, file = "fullmoon.csv", quote = FALSE, row.names = FALSE)
fullmoon <- read.csv("~/git_repositories/EPIB607/exercises/inferencemeans/fullmoon.csv")
fullmoon

moon.diff <- fullmoon$moon_days - fullmoon$other_days

moon.mean <- mean(moon.diff)
moon.sdev <- sd(moon.diff)
moon.n <- length(moon.diff)

t95.moon <- qt(0.975, df = moon.n-1)

moon.error <- t95.moon*moon.sdev/sqrt(moon.n)

CI95.moon.t <- moon.mean + c(-moon.error, moon.error)

#fullmoon$diff <- fullmoon$moon_days - fullmoon$other_days
# s_dist <- do(10000) * mean(~ diff, data = resample(fullmoon))
# hist(s_dist$mean)
# knitr::kable(fullmoon)
```

a. Calculate a 95% confidence interval for the mean difference in disruptive behaviors. State the assumptions you used to calculate this interval.
b. Test the hypothesis that moon days will lead to more disruptive behavior. State your assumptions and provide a brief conclusion based on your analysis. 
c. Find the minimum value of the mean difference in disruptive behaviors ($\bar{y}$) needed to reject the null hypothesis. 
d. What is the probability of detecting an increase of 1.0 aggressive behavior per day during moon days? 



