---
title: Logistic Regression in R

# Use letters for affiliations
author:
  - name: EPIB607 - Inferential Statistics
    affiliation: a
#  - name: Second Author
#    affiliation: a,b
address:
  - code: a
    address: Fall 2020, McGill University
#  - code: b
#    address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Bhatnagar

# Place DOI URL or CRAN Package URL here
#doi: "https://sahirbhatnagar.com/EPIB607/"

# Abstract
#abstract: |
#  This DALITE quiz will cover regression.

# Optional: Acknowledgements
#acknowledgements: |
#  This template package builds upon, and extends, the work of the excellent
#  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
#  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
#  gratefully acknowledged as this work would not have been possible without them.  Our extensions
#  are under the same respective licensing term
#  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
#  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).

# Optional: One or more keywords
keywords:
#  - linear regression
#  - poisson regression
#  - logistic regression

# Paper size for the document, values of letterpaper and a4paper
papersize: letter

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 12pt

# Optional: Force one-column layout, default is two-column
one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography 
bibliography: pinp

# Optional: Enable a 'Draft' watermark on the document
watermark: false

# Customize footer, eg by referencing the vignette
#footer_contents: "DALITE Q5 due October , 2019 by 5pm"

# Produce a pinp document
output: pinp::pinp

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{YourPackage-vignetteentry}
  %\VignetteKeywords{YourPackage, r, anotherkeyword}
  %\VignettePackage{Due Sepetember 26, 2018}
  %\VignetteEngine{knitr::rmarkdown}
---


# Fitting a Logistic Regression Model with a Single Predictor

The \textbf{\texttt{glm()}} function is used to fit logistic regression models. It has the following generic structure:

```{r, eval = FALSE}
glm(y ~ x, data, family = binomial(link = "logit"))
```

where the first argument specifies the variables used in the model; in this example, the model regresses a response variable \texttt{y} against an explanatory variable \texttt{x}. The second argument is used only when the dataframe name is not already specified in the first argument. Running the function creates an *object* (of class '\texttt{lm}' and '\texttt{glm}') that contains several components, such as the model coefficients. The model coefficients are directly displayed upon running \texttt{glm()}, while other components can be accessed through either the \texttt{\$} notation or specific functions like \texttt{summary()}. The argument \texttt{family = binomial(link = "logit")} is specific to logistic regression; the texttt{glm()} function is capable of running families of general linear models that are not discussed in this course.

The following example shows fitting a linear model that predicts the estimated log odds of death before discharge from resting heart rate, using data from \texttt{icu}.

```{r}
#load the data
library(aplore3)
data("icu")

#fitting logistic model
glm(sta ~ hra, data = icu, family = binomial(link = "logit"))
```

To fit a linear model that predicts the estimated log odds of survival to discharge from resting heart rate, it is necessary to relevel the factor \texttt{sta} such that a \texttt{1} corresponds to individuals who survived to discharge. This can be accomplished with \textbf{\texttt{factor()}} and \textbf{\texttt{rev()}}. The \texttt{rev()} function reverses elements. In the example below, applying \texttt{rev)} to a vector {1, 2, 3} produces a vector {3, 2, 1}.

```{r}
#check levels
levels(icu$sta)

#relevel survival
icu$sta = factor(icu$sta, levels = rev(levels(icu$sta)))

#check levels
levels(icu$sta)

#example of using rev()
a = c(1, 2, 3)
rev(a)
```

The following example shows outputting the model summary, selectively outputting model coefficients from the model fit, and extracting the numeric value of a coefficient.

```{r}
#name the model
model.hra = glm(sta ~ hra, data = icu, family = binomial(link = "logit"))

#model summary
summary(model.hra)

#model summary of coefficients
summary(model.hra)$coef

#extract value of slope coefficient
coef(model.hra)[2]
```

As in linear regression, the \texttt{predict()} function can be used to evaluate the regression equation for specific values of a predictor variable. The following example shows predicting the estimated log odds of survival to discharge for an individual with resting heart rate of 98 bpm.

```{r}
predict(model.hra, newdata = data.frame(hra = 98))
```


\newpage

# Multiple Logistic Regression

## Working with Several Predictors

The \textbf{\texttt{glm()}} function is used to fit linear models. It has the following generic structure:

```{r, eval = FALSE}
glm(y ~ x1 + x2, data, family = binomial(link = "logit"))
```

where the first argument specifies the variables used in the model; in this example, the model regresses a response variable \texttt{y} against two explanatory variables \texttt{x1} and \texttt{x2}. Additional predictor variables can be added to the model formula with the \texttt{+} symbol, and an interaction between two variables is specified with the \texttt{*} symbol.

The following example shows fitting a linear model that predicts the estimated log odds of survival to discharge from age and gender, and a linear model that predicts the estimated log odds of survival to discharge from age, gender, and their interaction.

```{r}
#fitting model with age and gender
glm(sta ~ age + gender, data = icu, family = binomial(link = "logit"))

#fitting model with age, gender, and an interaction term
glm(sta ~ age*gender, data = icu, family = binomial(link = "logit"))
```

## Calculating AIC

The AIC of a logistic model can be extracted from \texttt{summary()} or computed via the \textbf{\texttt{AIC()}} function.

The following example shows how to output the AIC from the model predicting estimated odds of survival to discharge from resting heart rate.

```{r}
#use summary()$aic
summary(model.hra)$aic

#use AIC()
AIC(model.hra)
```


## Collapsing Factor Levels

The \texttt{factor()} function can also be used to collapse levels of a factor. 

The following example shows the re-defining of the levels of \texttt{loc}; the variable initially has three levels (\texttt{Nothing}, \texttt{Stupor}, and \texttt{Coma}). The levels \texttt{Stupor} and \texttt{Coma} can be combined into a single level \texttt{Unconscious}, while the level \texttt{Nothing} is renamed \texttt{Conscious}. 

```{r}
#view levels of loc
levels(icu$loc)

#create the loc.binary variable
icu$loc.binary = icu$loc

#redefine the factor levels of loc.binary
levels(icu$loc.binary) = list("Conscious" = "Nothing", 
                              "Unconscious" = c("Stupor", "Coma"))

#view levels of loc.binary
levels(icu$loc.binary)

#compare tables
table(icu$loc); table(icu$loc.binary)
```

